% ══════════════════════════════════════════════════════════════
%  colophon.tex — Production notes
% ══════════════════════════════════════════════════════════════
\chapter*{Colophon}
\addcontentsline{toc}{chapter}{Colophon}

\begin{center}
  {\color{brand}\rule{0.4\textwidth}{1.5pt}}
\end{center}

\vspace{12pt}

\noindent
This book was typeset using \LaTeX{} with the \texttt{memoir}
document class, configured for Crown Quarto format
(7.44\,\texttimes\,9.69 inches).

\medskip
\noindent\textbf{Typography}

\smallskip
\noindent
The body text is set in \emph{TeX Gyre Pagella} (a Palatino
revival) at 11\,pt with 1.05\texttimes{} leading. Section
headings and navigational elements use \emph{TeX Gyre Heros}
(Helvetica family). Code listings and inline code use
\emph{TeX Gyre Cursor} (Courier family). Mathematical
notation uses \emph{Palatino Math} via the \texttt{mathpazo}
package. Microtypographic refinements---character protrusion,
font expansion, and tracking---are provided by the
\texttt{microtype} package.

\medskip
\noindent\textbf{Color system}

\smallskip
\noindent
The book's visual identity is built around Rust Orange
(\texttt{\#B7410E}) for accent elements and Dark Teal
(\texttt{\#2C5F7C}) for structural headings. Eleven callout box
families provide instant visual classification of notes, tips,
warnings, checkpoints, and code references.

\medskip
\noindent\textbf{Code presentation}

\smallskip
\noindent
Source code is syntax-highlighted using the \texttt{minted}
package with Pygments. Rust listings include line numbers and
support inline \LaTeX{} annotations via escape sequences.
Manning-style numbered annotation circles connect code lines
to explanatory notes below each listing.

\medskip
\noindent\textbf{Production tools}

\smallskip
\noindent
The PDF was produced with pdf\LaTeX{} and \texttt{latexmk}.
Diagrams were created with Ti\textit{k}Z. The index was generated
with \texttt{makeindex} and the glossary with
\texttt{glossaries-extra}. Version control was managed with Git.

\vfill
\begin{center}
  {\color{brand}\rule{0.4\textwidth}{1.5pt}}\\[8pt]
  {\small\sffamily\color{sectioncolor} First Edition\enspace---\enspace\the\year}
\end{center}
